Clustering based on probability distributions with application on residential customers

### 2021-09-29 main.Rmd
methodology first para re-write

---
title: Clustering of residential Australian smart meter customers

@@

@@

@@

## Missing Data

@@

## Raw data

@@

```{r load-elec, message = FALSE, warning=FALSE}

@@

```{r vismiss, eval = FALSE}
## transform all columns
customer_data <- customer_data %>% mutate_each(funs(empty_as_na))

vis_miss(customer_data, warn_large_data = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, size = 5))
```

```{r miss-var, eval = FALSE}
gg_miss_var(customer_data) + theme_bw() +
  theme(axis.text.y = element_text(size = 4))
```

```{r}
elec_ts <- read_rds("data/elec_ts.rds")
```

```{r raw-50, fig.height = 8, fig.cap="The raw data for 50 households are shown. It looks like there is a lot of missing values and unequal length of time series along with asynchronous periods for which data is observed."}

@@

## Common load clustering techniques of smart meter data

@@

## The proposed load-clustering technique

The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the probability distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a household and the underlying variable as the electricity demand. So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand.

@@

@@

<!-- Trend and seasonality are common features of time series, and it is natural to characterize a time series by its degree of trend and seasonality. By considering the probability distributions through the use of $wpd_{norm_{s,t}}(A,B)$, these features of the time series are lost and hence there is no need to de-trend or de-seasonalize the data before performing the clustering algorithm. -->

@@

```{r,warning=FALSE, message=FALSE,eval = FALSE}
library(readr)
library(here)
library(dplyr)
library(tsibble)
library(gravitas)
library(purrr)

set.seed(7777)
all_data <- read_rds(here("../paper-hakear/paper/sim_table/all_data_scalar.rds"))
sm_cust_data <- read_rds(here("../paper-hakear/paper/data/sm_cust_data.rds"))

sm_list <- sm_cust_data %>%
  filter(customer_id == "10006414") %>% 
  create_gran("day_week") %>% 
  tibble() %>% 
  ungroup() %>% 
  group_split(day_week)

p <- seq(0.01, 0.99,0.01)

dist_list <- purrr::map(sm_list, function(x){
    map(p, function(y){
       quantile(x$general_supply_kwh, probs = y)
    }) %>% unlist()
}) %>% bind_rows(.id = "gran_id")
# 
# lag <- dist_list[2,][-1] %>% t() %>% 
#   tibble() %>% 
#   lag() %>% acf()

```{r mds, warning=FALSE, message=FALSE, fig.cap="Multidimensional scaling of the 100 households is shown along with the groups formed by the hierarchical clustering.", eval = FALSE}
set.seed(7777)
sm_unique_cust <- sm_cust_data %>% distinct(customer_id)
sm_cust_samp <- sm_cust_data %>% 
  #filter(customer_id %in% sm_30) %>% # for demo show 30 customers
  #filter(year(reading_datetime) == 2012) %>% #for demo just show one column
  as_tsibble(index = reading_datetime,
             key = customer_id) %>% 
  create_gran("day_week") %>% 
  create_gran("hour_day")

#all_data <- write_rds(all_data, "paper/sim_table/all_data.rds")

facet_variable = "day_week"
x_variable = "hour_day"

one_harmony_data <- all_data %>%
  #filter(customer_id %in% sm_30) %>% # for demo show 30 customers
  filter(facet_variable == "day_week",
         x_variable == "hour_day") %>% 
  mutate(customer_id = as.character(customer_id))

mds<- one_harmony_data %>%
  select(customer_id, facet_variable, x_variable, wpd_norm) %>% 
  dist() %>%          
  cmdscale() %>%
  as_tibble()

colnames(mds) <- c("Dim.1", "Dim.2")
rownames(mds) <- unique(one_harmony_data$customer_id)
mds <- mds %>%
  mutate(customer_id = as.character(unique(one_harmony_data$customer_id))) %>%
  select((customer_id)
         , Dim.1, Dim.2)

# ggpubr::ggscatter(mds,
#                   x = "Dim.1",
#                   y = "Dim.2", 
#                   label = rownames(mds),
#                   size = 1,
#                   repel = TRUE)

# clustering as well

d = one_harmony_data %>%
  select(customer_id, facet_variable, x_variable, wpd_norm) %>% 
  dist()

hc = stats::hclust(d,method="complete")
#plot(hc)
groups<-cutree(hc, k=3)

all_data_cluster <- cbind(one_harmony_data, groups) %>% 
  left_join(mds, by = "customer_id") %>% 
  mutate(groups = as.factor(groups))

# q2 <- ggpubr::ggscatter(all_data_cluster,
#                   x = "Dim.1",
#                   y = "Dim.2", 
#                   label = rownames(mds),
#                   color = "groups",
#                   size = 1,
#                   repel = TRUE)

q2 <- all_data_cluster %>% 
  ggplot() +
  geom_point(aes(x = Dim.1, y = Dim.2, color = groups)) +
  scale_color_viridis_d()

 q2
```

```{r linear-cluster, eval = FALSE}
group1 <- all_data_cluster %>% 
  filter(groups == 1)

group2 <- all_data_cluster %>% 
  filter(groups == 2)

group3 <- all_data_cluster %>% 
  filter(groups == 3)

g1 <- sm_cust_samp %>% 
  filter(customer_id %in% group1$customer_id) %>% 
   ggplot(aes(x=reading_datetime, 
             y= general_supply_kwh), alpha = 0.5) + 
  geom_line() +
  scale_fill_gradient2() +
  facet_wrap(~customer_id, nrow = 1) +
  theme(strip.text.x = element_blank()) +
  theme_void() +
  ylab("group 1")

g2 <- sm_cust_samp %>% 
  filter(customer_id %in% group2$customer_id) %>% 
   ggplot(aes(x=reading_datetime, 
             y= general_supply_kwh), alpha = 0.5) + 
  geom_line() +
  scale_fill_gradient2() +
  facet_wrap(~customer_id, nrow= 1) +
  theme(strip.text.x = element_blank()) +
  theme_void() +
  ylab("group 2")

g3 <- sm_cust_samp %>% 
  filter(customer_id %in% group3$customer_id) %>% 
   ggplot(aes(x=reading_datetime, 
             y= general_supply_kwh), alpha = 0.5) + 
  geom_line() +
  scale_fill_gradient2() +
  theme(strip.text.x = element_blank()) +
  theme_void() +
  ylab("group 3")

# g1/g2 + g3

```{r cyclic-cluster, warnings = FALSE, message = FALSE, fig.cap = "Distribution of energy demand across day-of-week and hour-of-day for three groups clustered through the proposed methodology. For group-1, the morning peak looks very sharp, for group-2 the morning and evening peaks are rounded and spread over few hours, for group-3 peaks are less rounded. They differ in their relationship between 75th and 90th percentile.", eval = FALSE}
cg1 <- sm_cust_samp %>% 
  filter(customer_id %in% group1$customer_id) %>% 
  prob_plot("day_week", "hour_day", response = "general_supply_kwh",
            plot_type = "quantile", symmetric = TRUE, quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) +
  scale_color_brewer(palette = "RdPu") + ggtitle("group 1") +
  scale_x_discrete(breaks = seq(0, 23, 5))

cg2 <- sm_cust_samp %>% 
  filter(customer_id %in% group2$customer_id) %>% 
  prob_plot("day_week", "hour_day", response = "general_supply_kwh",
            plot_type = "quantile", symmetric = TRUE, quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) +
  scale_color_brewer(palette = "RdPu") + ggtitle("group 2") +
  scale_x_discrete(breaks = seq(0, 23, 5))

cg3 <- sm_cust_samp %>% 
  filter(customer_id %in% group3$customer_id) %>% 
  prob_plot("day_week", "hour_day", response = "general_supply_kwh",
            plot_type = "quantile", symmetric = TRUE, quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) +
  scale_color_brewer(palette = "RdPu") + ggtitle("group 3") +
  scale_x_discrete(breaks = seq(0, 23, 5))

cg1 
cg2
cg3

###  Distinction, repeatability, and robustness metrics

@@

## Cluster validation

@@

@@

## Conclusion

@@

# R packages

The implementation for our framework for the first paper is available in the R package [`gravitas`](https://github.com/Sayani07/gravitas) and that for the second paper is made available in the package [`hakear`](https://github.com/Sayani07/hakear) for ease of use in other applications.

# Discussion

The final chapter of my thesis will consist of the concluding remarks of the whole thesis and possible future extensions.

>>>>

---
title:  Clustering based on probability distributions with application on residential customers

@@

_Themes_

_Common load clustering techniques of smart meter data_

@@

### Data source

@@

### Raw data

### Missing Data

@@

13735 customers in elec_ts
8685 customers in elec_nogap
5050 customers in count_na_df

Then is the graph of missing observations even interesting.
You can show two graphs, one to show that missingness do not have a pattern
another to show even if no missing, they start and end at different times. (A sample of 50 customers).

## LGA and weather data

@@

@@

#NEW

The raw smart meter data is indexed by time for each key variable and includes values of different variables of interest (measurement variables) at each time point. However, this series could be represented in several possible ways. One sequence could represent distribution of hourly consumption across a single day, while another could track consumption across days of a week or months of a year. Depending on the cyclic granularity considered, each sequence could be thought of as a different data generating processes.

We can then model different data generation processes depending on our level of aggregation and time scale.

@@

@@

The data set solely contains readings from smart meters and no information about the consumers' specific physical, geographical, or behavioural attributes. As a result, no attempt is made to explain why consumption varies. Instead, this work investigates how much energy usage heterogeneity can be found in smart meter data and what some of the most common electricity use patterns are. It is worth noting that when studying these dynamics, a variety of objectives may be pursued. One objective could be to group consumers with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. Most clustering algorithms offer only daily energy profiles throughout the hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar shapes across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification.

The foundation of our method is unsupervised clustering algorithms based exclusively on time-series features. First, we study the underlying distributions that may have resulted in different patterns across temporal granularities in order to identify a mechanism to classify them based on the similarity of those distributions. Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this.

This section presents the work flow of the methodology:

- _Find significant harmonies or cyclic granularities_:

For each key variable, the raw smart meter data is indexed by time and comprises values of several variables of interest (measurement variables) at each time point. This sequence, though, could be depicted in a variety of ways. One sequence could reflect the distribution of hourly consumption over a single day, while another could indicate consumption over a week or a year. The sequence of measured variables throughout each category of any cyclic granularity could be regarded as a single data generating process. It is beneficial to evaluate those cyclic granularities that are significant for the majority of customers or noteworthy in an electricity-behavior context, so that when the customers are grouped, we can expect to observe some interesting patterns across the categories of the cyclic granularities considered.

- _Distance metric_

- _Clustering algorithm_

- _Characterization of clusters_

- A random sample  of the original data is taken for clustering analysis and includes missing and noisy observations (detailed description in Appendix)

- All harmonies are computed for each customer in the sample.
Cyclic granularities which are clashes for all customers in the sample are removed.

- It is worth noting that a number of other solutions may be considered at the pre-processing stage of the method. We have considered a) Normal-Quantile Transform and b) Robust transformation.

- Two methods are considered for computing dissimilarity between two customers. The first one involves computing according to one granularity is computed as the sum of the JS distances between distribution of all the categories of the granularity. When we consider more than one granularity, we consider the sum of the  average distances for all the granularity so that the combined metric is also a distance.

- Given the scale of dissimilarity among the energy readings, the model chooses optimal number of clusters

- Once clusters have been allocated, the groups are explored visually.

- Results are reported and compared.

Two methods are used for computing distances between subjects and then hierarchical clustering algorithm is used.

The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the probability distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a customer and the underlying variable as the electricity demand. So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand.

@@

### Notations

Consider an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the probability distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. In our work, we are using $i$ as denoting a household and the underlying variable as the electricity demand. Further consider a cyclic granularity of the form $B = \{ b_k: k = 1, 2, \dots, K\}$. Each customer consists of collection of probability distributions.

So $f_i(v)$ is the distribution of household $i$ and $v$ is electricity demand. We want to cluster distributions of the form $f_{i,j,k}(v)$, where $i$ and $j$ denote $i^{th}$ and $j^{th}$ customer respectively.

a harmony table consisting of many harmonies, each of the form $(A, B)$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$.

### A single or pair of granularities together (change names)

The methodology can be summarized in the following steps:

- _Pre-processing step_

Robust scaling method or NQT used for each customer.

- _NQT_

- _Treatment to outliers_

- _Handling trend, seasonality, non-stationarity and auto-correlation_
Trend and seasonality are common features of time series, and it is natural to characterize a time series by its degree of trend and seasonality. By considering the probability distributions through the use of $wpd_{norm_{s,t}}(A,B)$, these features of the time series are lost and hence there is no need to de-trend or de-seasonalize the data before performing the clustering algorithm. No need to exclude holiday or weekend patterns.

### Many granularities together (change names)

The methodology can be summarized in the following steps:

1. Compute quantiles of distributions across each category of the cyclic granularity
2. Compute JS distance between households for each each category of the cyclic granularity
3. Total distance between households computed as sum of JS distances for all hours
4. Cluster using this distance with hierarchical clustering algorithm (method "Ward.D")

_Pro:_  
- distance metric makes sense to group different shapes together  
- simulation results look great on typical designs  
_Cons:_  
- Can only take one granularity at once  
- Clustering a big blob of points together whereas the aim is to groups these big blob into smaller ones

### Multiple-granularities

_Description:_

Choose all significant granularities and compute wpd for all these granularities for all customers. Distance between customers is taken as the euclidean distances between them with the granularities being the variables and wpd being the value under each variable for which Euclidean distance needs to be measured.  
_Pro:_  
- Can only take many granularities at once
- can apply variable selection PCP and other interesting clustering techniques
- simulation results look great on typical designs
- splitting the data into similar sized groups  
_Cons:_  
- distance metric does not make sense to split the data into similar shaped clusters

@@

@@

_Distinction, repeatability, and robustness metrics_

@@

_Cluster validation_

@@

## Results

## Software implementation

The implementation for our framework is available in the R package [`gracsr`](https://github.com/Sayani07/gracsr) for ease of use in other applications.

# Results

## Clustering results for 100 customers

## Clustering results for 5K customers

@@

## Discussion

@@
